Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting

### 2023-08-24 doc/JAIR/jair-paper.tex

\newcommand{\makenode}[1]{\mathbf{#1}}
\newcommand{\nodeu}{\makenode{u}}
\newcommand{\nodev}{\makenode{v}}
\newcommand{\nodes}{\makenode{s}}
\newcommand{\nodep}{\makenode{p}}
\newcommand{\noder}{\makenode{r}}

@@

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\cdfour}{\progname{CD4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}
\newcommand{\Dtrim}{\progname{Drat-trim}}

@@

In more detail, \cdfour{} emits a sequence of clauses $R$ that
includes the conflict clauses that arose during a top-down processing of
the input clauses.  Given input formula $\phi_I$, their first task is to
check whether $\phi_I \imply R$, i.e., that any assignment that satisfies $\phi_I$ also satisfies each of the clauses in $R$.
They then base other parts of their proof on that
property and use a separate program to perform a series of additional
checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
satisifies the \emph{reverse asymmetric tautology} (RAT) property with
respect to the preceding
clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
clause $C$ to a set of clauses maintains satisfiability, but it does not necessarily
preserve models.  As an example, consider the following formulas:
\begin{center}
  \begin{tabular}{lccc}
    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_2 \lor \obar{x}_3)$\\
  \end{tabular}
\end{center}
Clearly, these two formulas are not equivalent---$\phi_1$ has six
models, while $\phi_2$ has four.  In particular, $\phi_1$ allows
arbitrary assignments to variable $x_2$.  Critically, however, the
second clause of $\phi_2$ is RAT with respect to $\phi_1$---any
satisfying assignment to $\phi_1$ can be transformed into one that
also satisfies $\phi_2$ by setting $x_2$ to 1, while keeping the values for other variables fixed.

@@

Each input clause deletion can be done as a single RUP step, based
on an algorithm to test for clausal entailment in det-DNNF graphs~\cite{capelli:sat:2019,darwiche:jair:2002}.  The
proof generator constructs the hint sequence from the defining
clauses of the POG nodes via a single, bottom-up pass through the
graph.  The RUP deletion proof for input clause $C$ effectively proves that any
total assignment $\assign$ that does not
satisfy $C$ will extend to assignment $\eassign$ such that
$\eassign(r) = 0$.  It starts with the set of literals
$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
assignment $\assign$ to falsify clause $C$.
It then
adds literals via unit propagation until a
conflict arises.    Unit literal $r$ gets
added right away, setting up a potential conflict.

@@

We set out to formally verify the system with two goals in mind: first, to ensure that the CPOG framework is mathematically
sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \ref{fig:chain}). These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}. Verification was greatly aided by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean} automated tactic. In this section, we briefly describe the functionality we implemented and what we proved about it. More information is provided in the supplement~\cite{bryant:sat:2023:supplement}.

@@

\section{Implementations}
We have implemented programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}.\footnote{The source code for all tools
is available at \url{https://github.com/rebryant/cpog/releases/tag/v1.0.0}.}  The proof generator is the same in both
cases, since it need not be trusted.
Our \emph{verified}
version of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our \emph{prototype} version is written in C\@.
It is faster and
more scalable, but we anticipate its need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

Our proof generator is written in C/C++ and uses
\cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert
proof steps back into hinted CPOG clause additions, the generator can
use either its own RUP proof generator, or it can invoke
\dtrim{}~\cite{heule:cade:2013}.  The latter yields shorter proofs and scales well
to large proofs, but each invocation has a high startup cost.  We
therefore only use it when solving larger problems (currently ones
with over 1000 clauses).

@@

We incorporated a ring evaluator into the prototype checker.  It
can perform both standard and weighted
model counting with full precision.  It performs arithmetic over a subset of the rationals
we call $\drational$, consisting of numbers of the form $a \cdot 2^{b}
\cdot 5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented
to have arbitrary range.  Allowing scaling by powers of 2 enables the
density computation and rescaling required for standard model
counting.  Allowing scaling by powers of both 2 and 5 enables exact
decimal arithmetic, handling the weights used in the weighted model
counting competitions.  To give a sense of scale, the counter
generated a result with 260,909 decimal digits
for one of the weighted benchmarks.

@@

We are hopeful that having checkable proofs for knowledge compilers
will allow them to be used in applications where high levels of trust
are required, and that it will provide a useful tool for developers of
knowledge compilers.
%%Although our current implementation only handles
%%the outputs of one particular program, it could be adapted
Our experiments demonstrate that our toolchain can already handle
problems nearly at the limits of current knowledge compilers.  Further
engineering and optimization of our proof generator and checker could
improve their performance and capacity substantially.  We are hopeful
that our tool can be adapted to handle other knowledge compiler
representations, such as sentential decision diagrams (SDDs)~\cite{darwiche:ijcai:2011}.

\subsection{Generalizing to Semirings}

@@

%% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
%% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{1 \leq i \leq k}
%% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
%% such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
%% \cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
%% such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
%% \cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
%% of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
%% for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
%% the sum of $t_1, \ldots, t_k$.

>>>>

\newcommand{\makenode}[1]{\mathbf{#1}}
\newcommand{\nodeu}{\makenode{u}}
\newcommand{\nodev}{\makenode{v}}
\newcommand{\nodew}{\makenode{w}}
\newcommand{\nodes}{\makenode{s}}
\newcommand{\nodep}{\makenode{p}}
\newcommand{\noder}{\makenode{r}}

@@

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\cdfour}{\progname{CD4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}
\newcommand{\ltrim}{\progname{lrat-trim}}
\newcommand{\Dtrim}{\progname{Drat-trim}}

@@

In more detail, \cdfour{} emits a sequence of clauses $R$ that
includes the conflict clauses that arose during a top-down processing of
the input clauses.  Given input formula $\phi_I$, their first task is to
check whether $\phi_I \imply R$, i.e., that any assignment that satisfies $\phi_I$ also satisfies each of the clauses in $R$.
They then base other parts of their proof on that
property and use a separate program to perform a series of additional
checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
satisifies the \emph{reverse asymmetric tautology} (RAT) property with
respect to the preceding
clauses~\cite{jarvisalo:ijcar:2012,heule:cade:2013}.  Adding a RAT
clause $C$ to a set of clauses maintains satisfiability, but it does not necessarily
preserve models.  As an example, consider the following formulas:
\begin{center}
  \begin{tabular}{lccc}
    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_2 \lor \obar{x}_3)$\\
  \end{tabular}
\end{center}
Clearly, these two formulas are not equivalent---$\phi_1$ has six
models, while $\phi_2$ has four.  In particular, $\phi_1$ allows
arbitrary assignments to variable $x_2$.  Critically, however, the
second clause of $\phi_2$ is RAT with respect to $\phi_1$---any
satisfying assignment to $\phi_1$ can be transformed into one that
also satisfies $\phi_2$ by setting $x_2$ to 1, while keeping the values for other variables fixed.

@@

Each input clause deletion can be done as a single RUP step, based
on an algorithm to test for clausal entailment in det-DNNF graphs~\cite{darwiche:jair:2002,capelli:sat:2019}.  The
proof generator constructs the hint sequence from the defining
clauses of the POG nodes via a single, bottom-up pass through the
graph.  The RUP deletion proof for input clause $C$ effectively proves that any
total assignment $\assign$ that does not
satisfy $C$ will extend to assignment $\eassign$ such that
$\eassign(r) = 0$.  It starts with the set of literals
$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
assignment $\assign$ to falsify clause $C$.
It then
adds literals via unit propagation until a
conflict arises.    Unit literal $r$ gets
added right away, setting up a potential conflict.

@@

We set out to formally verify the system with two goals in mind: first, to ensure that the CPOG framework is mathematically
sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \ref{fig:chain}). These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}. Verification was greatly aided by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean} automated tactic. In this section, we briefly describe the functionality we implemented and what we proved about it.

@@

\section{Implementation}
We have implemented programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}.\footnote{The source code for all tools
is available at \url{https://github.com/rebryant/cpog/releases/tag/v1.0.0}.}  The proof generator is the same in both
cases, since it need not be trusted.
Our \emph{verified}
version of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our \emph{prototype} version is written in C\@.
It is faster and
more scalable, but we anticipate its need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

Our proof generator is written in C/C++ and uses a recent version of
the \cadical{} SAT solver that directly generates hinted proofs in
LRAT format~\cite{biere:sat:2023}.  It also uses their tool \ltrim{}
to reduce the length of the generated proofs.

@@

We incorporated a ring evaluator into the prototype checker.  It can
perform both standard and weighted model counting with full precision.
It performs arithmetic over a subset of the rationals we call
$\drational$, consisting of numbers of the form $a \cdot 2^{b} \cdot
5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented to
have arbitrary range.  Allowing scaling by powers of 2 enables the
density computation and rescaling required for standard model
counting.  Allowing scaling by powers of both 2 and 5 enables exact
decimal arithmetic, handling the weights used in the weighted model
counting competitions.  To give a sense of scale, the counter
generated a result with 260,909 decimal digits for one of the weighted
benchmarks.  Our implementation of arbitrary-range integers represents
a number as a sequence of ``digits'' with each digit ranging from $0$
to $10^9-1$, and with the digits stored as four-byte blocks.  This
allows easy conversion to and from a decimal representation of the
number.

@@

\section{Extensions}

We are hopeful that having checkable proofs for knowledge compilers
will allow them to be used in applications where high levels of trust
are required, and that it will provide a useful tool for developers of
knowledge compilers.  Our current implementation only handles the
outputs of knowledge compilers that generate dec-DNNF graphs, and it
supports only queries that can be computed via ring evaluation.  Here
we discuss ways to extend both capabilities.

\subsection{Validating Sentential Decision Diagrams}
\label{sect:sdd}

\begin{figure}
  \begin{center}
    \input{dd/sdd}
  \end{center}
\caption{Translation of SDD node $\nodeu$ into POG node $\hat{\nodeu}$.  An SDD node with $2k$ children expands into $k$ product nodes and $k-1$ sum nodes.}
\label{fig:sdd:pog}
\end{figure}

\citeA{darwiche:ijcai:2011} introduced \emph{Sentential
Decision Diagrams} (SDDs) as a generalization of 
Ordered Binary Decision Diagrams (OBDDs)~\cite{Bryant:1986}.  Like
OBDDs, SDDs are a canonical representation and support bottom-up
construction.

As with other representations, each node $\nodeu$ in an SDD denotes a
Boolean formula $\phi_{\nodeu}$.  Leaf nodes correspond to $\tautology$, $\nil$, and literals.
Each nonterminal node has $2k$ children,
organized as pairs $\nodep_i$ and $\nodes_i$ for $1\leq i \leq k$.
The dependency set for the node must be partitioned into disjoint sets $X_p$ and $X_s$ such that
$\dependencyset(\phi_{\nodep_i}) \subseteq X_p$ and
$\dependencyset(\phi_{\nodes_i}) \subseteq X_s$ for all $1 \leq i \leq k$.
Children $\nodep_1, \nodep_2, \ldots, \nodep_k$ serve as a set of
\emph{selector} functions with mutually disjoint models.  That is,
$\modelset(\phi_{\nodep_i}) \cap \modelset(\phi_{\nodep_j}) =
\emptyset$ for $1 \leq i < j \leq k$.
The Boolean formula associated with node $\nodeu$ is then
\begin{eqnarray}
\phi_{\nodeu} & = & \bigvee_{1 \leq i \leq k} \left[\phi_{\nodep_i} \land \phi_{\nodes_i}\right] \label{eqn:sdd}
\end{eqnarray}

An SDD can be translated into a POG, where an SDD node with $2k$
children expands into $k$ product nodes and $k-1$ sum nodes.
In particular, assume that each nonterminal SDD node $\nodev$ can be represented by a
subgraph in the POG with root node $\hat{\nodev}$.  For leaf node $\nodev$, its counterpart $\hat{\nodev}$ is
the associated constant or literal, rather than a POG node.  Figure~\ref{fig:sdd:pog} then
illustrates how node $\nodeu$ with $2k$ children would be expanded
into a subgraph with root node $\hat{\nodeu}$.  We can see that each
of the product nodes has arguments with disjoint dependency sets, and
each of the sum nodes has disjoint sets of models.

Several different methods could be used to generate a CPOG
certification from an SDD\@.  For the forward implication proof, the
monolithic approach we have used could generate the proof via a single
call to a proof-generating SAT solver.  For an SDD generated by the
bottom up approach originally proposed by \citeA{darwiche:ijcai:2011},
the generator could be modified to generate proof steps for each
recursive call during the construction, as is done to generate proofs
of unsatisfiability with proof-generating, BDD-based SAT
solvers~\cite{Jussila:2006,ebddres,bryant:fmcad:2022,bryant:tocl:2023}.
For an SDD generated by a top-down process, a structural method
similar to the one we have described could be applied.  A further
requirement would be to generate mutual exclusion proofs for the $k-1$
sum nodes illustrated in Figure~\ref{fig:sdd:pog}.  These would be
based on the disjoint model property of the $k$ nodes of the form
$\hat{\nodep}_i$.  One approach would be to simply make $k-1$ calls a
proof-generating SAT solver.

\subsection{Generalizing to Semirings}
\label{sect:semiring}

@@

Our experiments demonstrate that our toolchain can already handle
problems nearly at the limits of current knowledge compilers.  Further
engineering and optimization of our proof generator and checker could
improve their performance and capacity substantially.
